Magnetised Polish doughnuts revisited

### 2017-07-04 paper_revised/reply_referee.tex

\bigskip

{\bf R:} - Sec. 2.1: the authors introduce the concept of cusp but do not really
define it

\bigskip

{\bf A:} We have included the following definition: ``The cusp is defined as the circle 
in the equatorial plane on which the pressure gradient vanishes and the angular 
momentum of the disc equals the Keplerian angular momentum."

@@

\bigskip

{\bf R:} - Sec. 2.1: they introduce the equation of state (13) but do not discuss
how they treat the specific internal energy that is part of the
enthalpy used in (13). A comment should be added.

\bigskip

{\bf A:} In our work we are following the same procedure as in Komissarov (2006), where no comment on the treatment
of the specific internal energy is explicitly provided. Since we have been guided by this previous result, we have used the
same polytropic relationship between the enthalpy and the fluid pressure, Eq.~(12), and between the rest-mass density and the
fluid pressure (as can be read in the paragraph following Eq.~(27)). Thanks to the remark of the referee we now realise 
that by doing this we (and Komissarov) are making an implicit assumption about the internal energy. By looking at the 
expression of the stress-energy tensor in Eq.~(9) of the manuscript, we see that the fluid enthalpy $w$ includes the 
rest-mass density $\rho$, and can thus be defined as $w=\rho h$ where $h$ is the specific enthalpy. The relativistic definition 
of this quantity (in units with $c=1$) is $h=1+\varepsilon+p/\rho$, where $\varepsilon$ is the specific internal energy and 
$p$ is the fluid pressure. From a thermodynamical point of view a fluid is non-relativistic in the limit $h=1$. Therefore,
since both equations of state we use are polytropes (relating $p$ with $w$ or with $\rho$ in the same functional form) we 
conclude that we are implictly assuming that $h=1$ in our models, i.e.~the disks we build in our procedure are non-relativistic 
from a thermodynamical point of view.

@@

\bigskip

We have included the following comment at the end of the paragraph following Eq.~(27): 
``We note that the fluid enthalpy $w$ includes the rest-mass density $\rho$, and can thus be defined as $w=\rho h$ where
$h$ is the specific enthalpy. The relativistic definition of this quantity is $h=1+\varepsilon+p/\rho$, where $\varepsilon$ is
the specific internal energy. From a thermodynamical point of view a non-relativistic fluid satisfies $h=1$. Therefore,
since we use poytropic equations of state (relating $p$ with either $w$ or $\rho$ in the same functional form) we are
implictly assuming that $h=1$, i.e.~the discs we build in our procedure are non-relativistic from a thermodynamical point of view."

@@

\bigskip

We have incorporated this simple demonstration as Appendix A in the manuscript.

\bigskip

We have also incorporated a second appendix in the manuscript, Appendix B, to explain why indeed, as shown in the inset of the right panel of Fig.~6, all models cross at $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$ for a very specific value of $\beta_{\mathrm{m}_{\mathrm{c}}}$. Correspondingly, we have added the following text towards the end of Section 4: ``{It must be noted that the location of the maximum of the magnetic pressure is identical for all models considered when $\beta_{\mathrm{m}_{\mathrm{c}}} \equiv 1/(\lambda - 1) = 3$, as we show in Appendix~B. At this value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ all models cross at $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$, as it is more clearly shown in the inset of the right panel of Fig.~6.}"

@@

\bigskip

{\bf A:} We have built models with very small values of $\beta_{m_c}$, such as  $\beta_{m_c} = 10^{-20}$ (extremely high magnetisation) without encountering numerical difficulties. In practice we have found that there are no qualitative differences in the structure of the discs once $\beta_{m_c}$ becomes smaller than a sufficiently small value, about $\beta_{m_c}=10^{-3}$. This is
the reason why we choose that particular value as our lower limit in the figures and tables of the manuscript.

\bigskip

{\bf R:} - Sec.~4: I find the discussion in this section the less clear; a lot
of the considerations appear phenomenological but I can't find the
simple back-of-the-envelope demonstration that what the authors observe
numerically is what should be expected on physical grounds. For
example, the authors find that the $\rho_{\mathrm{max}}$ increases with magnetization
(left panel of Fig. 6) or that the position of the rest-mass density
maximum moves in with increasing magnetization (middle panel of
Fig. 6). What are the physical reasons for this behavior? I have some
suggestions on this but I urge the authors to provide some simple and
intuitive explanations first.

@@

\bigskip

{\bf A:} We appreciate the comment of the referee and apologize if the discussion is Section 4 is not sufficiently clear. We have been
analyzing in more detail our findings and have come up with the following simple explanations for our numerical observations.

To explain the movement of the maximum of the rest-mass density towards smaller $r$ we can start by computing the radial derivative of the enthaply, which should vanish at $r=r_{\rm max}$. From Eq.~(20) we have
\begin{eqnarray*}
\partial_r w=\partial_r\left[
-\Delta W \left( \frac{\lambda-1}{\lambda}\right)\frac{1}{K+K_m \mathcal{L}^{\lambda-1}}
\right]^{\frac{1}{\lambda-1}}\,,
\end{eqnarray*}
where we have used $\kappa=\lambda$ and have defined $\Delta W\equiv W-W_{\rm in}$. After some algebra, we arrive at the following expression for the radial derivative of the potential:
\begin{eqnarray*}
\partial_r W(r)|_{r=r_{\rm max}}=\frac{(\lambda-1) \Delta W(r_{\rm max})}{\mathcal{L}(r_{\rm max})}
\left[
\frac{(\partial_r \mathcal{L})|_{r=r_{\rm max}}}{1+\beta_{{\rm m}_{\rm c}}\left(\frac{\mathcal{L}(r_c)}{\mathcal{L}(r_{\rm max})} \right)^{\lambda-1}}
\right]\,.
\end{eqnarray*}
All the factors on the r.h.s.~of this expression are positive, except for $\Delta W(r_{\rm max})$ which is negative. For $\beta_{{\rm m}_{\rm c}}\rightarrow \infty$,  $\partial_r W(r)|_{r=r_{\rm max}}=0$, which is attained at $r_{\rm max}=r_c$, which is a minimum of the potential.
As $\beta_{{\rm m}_{\rm c}}$ decreases (and hence the magnetisation increases) the r.h.s.~increases in absolute value being always negative. Therefore, the position of $r_{\rm max}$ has to move towards smaller radii.

On the other hand, to explain the increase of $\rho_{\mathrm{max}}$ with magnetisation we can start with the same equation for the enthalpy $w$ as before (in our classical thermodynamical limit $w=\rho$). Next, we take $K=p_c$ and $K_m=\frac{p_c}{\beta_{{\rm m}_{\rm c}} \mathcal{L}_{c}^{\lambda-1}}$, where subindex $c$ means that the corresponding expressions are evaluated at the centre of the disc. We also use Eq.~(25) to write
\begin{eqnarray*}
p_c=-\Delta W_c \left( \frac{\lambda-1}{\lambda} \right) \left(  \frac{\beta_{{\rm m}_{\rm c}}}{1+\beta_{{\rm m}_{\rm c}}} \right)\,.
\end{eqnarray*}
Substituting these expressions in the equation for the enthalpy we obtain:
\begin{eqnarray*}
w|_{r=r_{\rm max}}= \left[
\frac{\Delta W(r_{\rm max})}{\Delta W(r_{\rm c})}\left(
\frac{1+\beta_{{\rm m}_{\rm c}}}{\left( \frac{\mathcal{L}(r_{\rm max})}{\mathcal{L}(r_c)}   \right)^{\lambda-1}+\beta_{{\rm m}_{\rm c}}}
\right)
\right]^{\frac{1}{\lambda-1}}\,.
\end{eqnarray*}
We note that $\frac{\mathcal{L}(r_{\rm max})}{\mathcal{L}(r_c)}\le 1$ and therefore the factor in parenthesis on the r.h.s.~is always $\ge 1$. Despite $\frac{\Delta W(r_{\rm max})}{\Delta W(r_{\rm c})}\le 1$ our numerical models are such that the product of the two factors on the r.h.s.~is always $\ge 1$. Therefore, the value of the rest-mass density at the maximum, $\rho_{\rm max}$, increases with magnetisation.

\bigskip
If the referee considered these explanations satisfactory and would like to have them included in the manuscript, we would be happy to do so.

### 2017-07-04 paper_revised/paper_revised.tex
work on the paper

\section{Value of the magnetisation parameter for $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$}\label{app_magmax}
In this appendix we derive the condition $\beta_{\mathrm{m}_{\mathrm{c}}} = 1/(\lambda - 1)$ for $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$.
First, we can use Eq.~\eqref{eq:eos_mag} to write 
\begin{eqnarray}\label{eq_max}
\frac{\partial{p_{\mathrm{m}}(r)}}{\partial r} = \partial_{r} \left(K_{\mathrm{m}} \mathcal{L}^{\lambda -1} w^{\lambda}\right) = 0\,,
\end{eqnarray}
which yields
\begin{eqnarray}
\partial_{r} p_{\mathrm{m}} = K_{\mathrm{m}} \mathcal{L}^{\lambda - 2}w^{\lambda} [(\lambda - 1)(\partial_r \mathcal{L}) w + \lambda \mathcal{L} (\partial_r w)]=0\,.
\end{eqnarray}
Inside the disc $K_{\mathrm{m}} \mathcal{L}^{\lambda - 2}w^{\lambda} \neq 0$. Therefore, to fulfill the extremum condition~\eqref{eq_max} we need 
\begin{equation}\label{eq:simple_expr}
(\lambda - 1)(\partial_r \mathcal{L}) w + \lambda \mathcal{L} (\partial_r w) = 0\,.
\end{equation}
To evaluate this expression, we need to compute the partial derivatives $\partial_r \mathcal{L}$ and $\partial_r w$. The derivative of $\mathcal{L}$ is straightforward, since $\mathcal{L} = \Delta \sin^2 \theta$,
\begin{equation}
\partial_r \mathcal{L} = 2(r-M)\sin^2 \theta\,.
\end{equation}
Let us now discuss $\partial_r w$. From Eq.~\eqref{eq:enthalpy_eq} we have
\begin{equation}
w = \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{\frac{1}{\lambda-1}}\,.
\end{equation}
Taking its derivative we obtain
\begin{eqnarray}
\partial_r w &=& \frac{w}{\lambda - 1} \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{-1} 
\nonumber \\
&\times&
\left[\left(-\partial_r W \left(\frac{\lambda -1 }{\lambda}\right) \frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right) + \left(-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{-(\lambda - 1) K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{(K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1})^2}\right)\right]\,.
\end{eqnarray}
Since at $r = r_{\mathrm{c}}$ the derivative of the potential is zero, $\partial_r W(r, \pi/2)|_{r = r_{\mathrm{c}}} = 0$, we can simplify the above expression to obtain
\begin{eqnarray}
\partial_r w &=& \frac{w}{\lambda - 1} \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{-1} 
\nonumber \\
&\times&
\left(-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right)\left[\frac{-(\lambda - 1) K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]\,,
\end{eqnarray}
which can be further simplified to the following form
\begin{equation}
\partial_r w = - w \left[\frac{K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]\,.
\end{equation}
Next, by inserting this expression in Eq.~\eqref{eq:simple_expr} we obtain
\begin{equation}
(\partial_r \mathcal{L}) w [(\lambda -1)(K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}) - \lambda K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}] = 0\,.
\end{equation}
Since $(\partial_r \mathcal{L}) w \neq 0$ inside the disc, the above equation leads to 
\begin{equation}
K (\lambda - 1) = K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}\,.
\end{equation}
Moreover, we can use Eqs.~\eqref{eq:eos_fluid}, \eqref{eq:eos_mag} and \eqref{eq:beta_eq} at the center (remember that $w_{\mathrm{c}} = 1$) to write $K$ as $K = \beta_{\mathrm{m}_{\mathrm{c}}} K_\mathrm{m}\mathcal{L}^{\lambda -1}$. Therefore, this allows us to obtain the following simple expression for the value of the magnetisation parameter at the center of the disc
\begin{equation}
\beta_{\mathrm{m}_{\mathrm{c}}} = \frac{1}{\lambda - 1}\,.
\end{equation}

\sg{We should add a short discussion of the generality of this result.}

>>>>

\section{Value of the magnetisation parameter for $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$}\label{app_magmax}
In this appendix we derive the condition $\beta_{\mathrm{m}_{\mathrm{c}}} = 1/(\lambda - 1)$ for $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$.
First, we can use Eq.~\eqref{eq:eos_mag} to write 
\begin{eqnarray}\label{eq_max}
\frac{\partial{p_{\mathrm{m}}(r)}}{\partial r} = \partial_{r} \left(K_{\mathrm{m}} \mathcal{L}^{\lambda -1} w^{\lambda}\right) = 0\,,
\end{eqnarray}
which yields
\begin{eqnarray}
\partial_{r} p_{\mathrm{m}} = K_{\mathrm{m}} \mathcal{L}^{\lambda - 2}w^{\lambda} [(\lambda - 1)(\partial_r \mathcal{L}) w + \lambda \mathcal{L} (\partial_r w)]=0\,.
\end{eqnarray}
Inside the disc $K_{\mathrm{m}} \mathcal{L}^{\lambda - 2}w^{\lambda} \neq 0$. Therefore, to fulfill the extremum condition~\eqref{eq_max} we need 
\begin{equation}\label{eq:simple_expr}
(\lambda - 1)(\partial_r \mathcal{L}) w + \lambda \mathcal{L} (\partial_r w) = 0\,.
\end{equation}
To evaluate this expression, we need to compute the partial derivatives $\partial_r \mathcal{L}$ and $\partial_r w$. The derivative of $\mathcal{L}$ is straightforward, since $\mathcal{L} = \Delta \sin^2 \theta$,
\begin{equation}
\partial_r \mathcal{L} = 2(r-M)\sin^2 \theta\,.
\end{equation}
Let us now discuss $\partial_r w$. From Eq.~\eqref{eq:enthalpy_eq} we have
\begin{equation}
w = \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{\frac{1}{\lambda-1}}\,.
\end{equation}
Taking its derivative we obtain
\begin{eqnarray}
\partial_r w &=& \frac{w}{\lambda - 1} \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{-1} 
\nonumber \\
&\times&
\left[\left(-\partial_r W \left(\frac{\lambda -1 }{\lambda}\right) \frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right) + \left(-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{-(\lambda - 1) K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{(K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1})^2}\right)\right]\,.
\end{eqnarray}
Since at $r = r_{\mathrm{c}}$ the derivative of the potential is zero, $\partial_r W(r, \pi/2)|_{r = r_{\mathrm{c}}} = 0$, we can simplify the above expression to obtain
\begin{eqnarray}
\partial_r w &=& \frac{w}{\lambda - 1} \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{-1} 
\nonumber \\
&\times&
\left(-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right)\left[\frac{-(\lambda - 1) K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]\,,
\end{eqnarray}
which can be further simplified to the following form
\begin{equation}
\partial_r w = - w \left[\frac{K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]\,.
\end{equation}
Next, by inserting this expression in Eq.~\eqref{eq:simple_expr} we obtain
\begin{equation}
(\partial_r \mathcal{L}) w [(\lambda -1)(K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}) - \lambda K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}] = 0\,.
\end{equation}
Since $(\partial_r \mathcal{L}) w \neq 0$ inside the disc, the above equation leads to 
\begin{equation}
K (\lambda - 1) = K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}\,.
\end{equation}
Moreover, we can use Eqs.~\eqref{eq:eos_fluid}, \eqref{eq:eos_mag} and \eqref{eq:beta_eq} at the center (remember that $w_{\mathrm{c}} = 1$) to write $K$ as $K = \beta_{\mathrm{m}_{\mathrm{c}}} K_\mathrm{m}\mathcal{L}^{\lambda -1}$. Therefore, this allows us to obtain the following simple expression for the value of the magnetisation parameter at the center of the disc
\begin{equation}
\beta_{\mathrm{m}_{\mathrm{c}}} = \frac{1}{\lambda - 1}\,.
\end{equation}
It is relevant to note that the only reference to the explicit form of the metric is done to show that $\mathcal{L} \neq 0$ and $\partial r \mathcal{L} \neq 0$ inside the disc. This means that our result holds for any stationary and axisymmetric metric if the aforementioned inequalities hold. Moreover, the result is also true for any angular momentum distribution that allows for the existence of a cusp and a centre (and $r_{\mathrm{c}} \neq r_{\mathrm{cusp}}$).
